---
title: "Wrangling Missing Data"
author: "Lucas Mendicino"
date: "3/3/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(naniar)
```



# Wrangling Missing Values

Ideally, a dataset would have NA for all missing values, but that is rarely the case. One must search for all missing values and replace them with NA.

Let's conduct this with the pacman dataset. 

```{r}
pacman <- read_csv("~/Desktop/pacman.txt")
pacman <- pacman %>% select(year, initial, score)
head(pacman)
```


The pacman dataset has pacman scores and we will focus on three columns: year, initial, and score

Let's use miss_scan_count to search for strange missing values of the form ("N/A", "missing", "na", " ")

We can do it indivudally like this:
```{r}
miss_scan_count(data = pacman, search = list("N/A"))
miss_scan_count(data = pacman, search = list("missing"))
miss_scan_count(data = pacman, search = list("na"))
miss_scan_count(data = pacman, search = list(" "))
```

Or all together:
```{r}
miss_scan_count(data = pacman, search = list("N/A", "missing","na", " "))
```

We can replace these values with NA using replace_with_na() and test again.

```{r}
pacman_clean <- replace_with_na(pacman, replace = list(year = c("N/A", "na", "missing"),
                                score = c("N/A", "na", "missing")))
                                        
miss_scan_count(pacman_clean, search = list("N/A", "na", "missing"))
```

We can also use the following functions on scoped variants. First, we just look at year using replace_with_na_at(), then only character variables using replace_with_na_if(), and finally all variables using replace_with_na_all().

```{r}
replace_with_na_at(pacman,
                   .vars = c("year"), 
                   ~.x %in% c("N/A", "missing", "na", " "))

replace_with_na_if(pacman,
                   .predicate = is.character, 
                   ~.x %in% c("N/A", "missing", "na", " "))

replace_with_na_all(pacman, ~.x %in% c("N/A", "missing", "na", " "))
```


## Handling implicit missing values

Sometimes a dataset has missing missing values, and it is implied that it is missing by its absence.

Let's explore this scenario using the frogger dataset, it contains scores for four players at four different times throughout the day.

```{r}
frogger <- read_csv("~/Desktop/Frogger.csv")
frogger <- frogger %>% select(name, time, value)
print(frogger)
```
We can see that andy, nic, and dan do not have four scores, they are missing scores from parts of the day and it is implied becuase they are not there.


Let's use functions from tidyr to fix implicit missing values.


complete() function makes implicit missing values explicit
```{r}
frogger_tidy <- frogger %>% complete(time, name)
print(frogger_tidy)
```


We can also use fill() to fill down the name variable in the dataset. It uses last observation carried down and only works for a specific structure.

```{r}
frogger_tidy %>% fill(name)
print(frogger_tidy)
```
Now, we see explicit missing values for andy, nic, and dan.

We can use fill() and complete() together

```{r}
frogger %>%
  fill(name) %>%
  complete(name, time)
```